---
title: "Introduction to Drifloon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Drifloon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

## Overview

The Drifloon package helps you download Canadian hourly weather station data 
from the Environment and Climate Change Canada portal.

Core functions:

- `load_metadata()`: Load station inventory used by all download wrappers.
- `sync_metadata()`: Checks and updates metadata from the official station inventory CSV.
- `download_station_month()`: Download one station, one year, one month.
- `download_station_by_name()`: Download one station for a year range.
- `download_station_province()`: Download all stations in one province.
- `download_all_station()`: Download all stations in your metadata.
- `download_metadata()`: Copy packaged metadata `.rds` to your output folder.

## Load metadata

```{r}
library(Drifloon)

meta <- load_metadata()
head(meta[, c("Name", "Province", "Station.ID", "HLY.First.Year", "HLY.Last.Year")])
```

The metadata frame is expected to include these key columns:

- `Name`: The name of the station. Note: some stations have the same name but different station ids.
- `Province`: The Province of the station. 
- `Station.ID`: The unique identifier for each station
- `HLY.First.Year`: The first year with available hourly data
- `HLY.Last.Year`: The last year with available hourly data

## Sync metadata from source inventory

Use `sync_metadata()` to pull the most recent station metadata and update local
metadata. This will only change when new station metadata is available.

```{r, eval = FALSE}
# Typical sync with defaults
sync_result <- sync_metadata()
str(sync_result)

# Example with tighter filtering and backup control
sync_result <- sync_metadata(
  provinces = c("ON", "QC"),
  min_year = 2000,
  create_backup = TRUE,
  max_backup_bytes = 25 * 1024^2
)
```

If you want this to happen during load, use:

```{r, eval = FALSE}
meta <- load_metadata(sync = TRUE)
```

## Download metadata snapshot

If you want a local copy of packaged metadata, use `download_metadata()`.

```{r, eval = FALSE}
metadata_path <- download_metadata(out_dir = "drifloon_output", overwrite = TRUE)
metadata_path
```

## Download a single month first

Start small to verify names, IDs, output paths. The default download directory is in the root of your current working directory in a folder `drifloon_output/`.

```{r, eval = FALSE}
# By station name
ok <- download_station_month(
  station_name = "TORONTO CITY",
  year = 2020,
  month = 1,
  out_dir = "drifloon_output"
)

# By station ID (useful when names are duplicated)
ok <- download_station_month(
  station_id = 31688,
  year = 2020,
  month = 1,
  out_dir = "drifloon_output"
)
```

## Download a station across a year range

Here we are asking for the station data for every month of the year range 2018 to
2020. Note you can use the `parallel=TRUE` argument if you want to parallelize 
download, which could make it faster. 

```{r, eval = FALSE}
download_station_by_name(
  station_name = "TORONTO CITY",
  first_year = 2018,
  last_year = 2020,
  out_dir = "drifloon_output",
  parallel = FALSE
)
```

There are multiple stations with the same name. In this case pass both 
`station_name` and `station_id` to force a specific match.

## Download by province

Because of how large the download of `download_station_province()` typically is,
it estimates total files and storage first. 

```{r, eval = FALSE}
# Interactive confirmation (default)
download_station_province(
  province = "ON",
  first_year = 2019,
  last_year = 2020,
  out_dir = "drifloon_output"
)

# Non-interactive workflows (scripts/CI)
download_station_province(
  province = "ON",
  first_year = 2019,
  last_year = 2020,
  out_dir = "drifloon_output",
  confirm = TRUE
)
```

## Download all stations

Use cautiously: this can generate very large downloads.

```{r, eval = FALSE}
download_all_station(
  first_year = 2019,
  last_year = 2020,
  out_dir = "drifloon_output",
  confirm = TRUE
)
```
## Create a SQL databse

The default directory is your current working directory but it can be specified ot be anything
you like. 

```{r, eval = FALSE}
create_database()
```

## Helper internals

- Metadata and paths:
  - `.load_station_metadata()`
  - `.resolve_out_dir()`
  - `.sync_station_metadata()`
- Name and province resolution:
  - `.normalize_station_name_key()`
  - `.find_station_name_index()`
  - `.province_normalize()`
- Validation and safety:
  - `.validate_year()`
  - `.estimate_download_size()`
  - `.check_disk_space()`
  - `.is_valid_downloaded_month_file()`

You can inspect internals for debugging with triple-colon access:

```{r, eval = FALSE}
Drifloon:::.normalize_station_name_key(c("Montréal/Trudeau", "Montreal Trudeau"))
Drifloon:::.province_normalize(c("on", "Ontario", "QC"))
```

## Typical end-to-end workflow

```{r, eval = FALSE}
library(Drifloon)

# 1) Refresh metadata
sync_metadata(provinces = c("ON", "QC"), min_year = 2000)

# 2) Load filtered metadata
meta <- load_metadata()

# 3) Validate one month first
download_station_month(
  station_name = "TORONTO CITY",
  year = 2020,
  month = 1,
  out_dir = "drifloon_output"
)

# 4) Scale to a station range
download_station_by_name(
  station_name = "TORONTO CITY",
  first_year = 2018,
  last_year = 2020,
  out_dir = "drifloon_output"
)

# 5) Scale to province only after checks
download_station_province(
  province = "ON",
  first_year = 2019,
  last_year = 2020,
  out_dir = "drifloon_output",
  confirm = TRUE
)

# 6) Create relational database
create_database()
```
